---
title: "Analysis"
author: "Yifeng Cao"
format: html
---

# **Level 1: Trends in Statistical Design (2010–2026)**

-   **Objective:** To quantify the adoption of Bayesian Adaptive Designs and Master Protocols in NSCLC.

-   **Methodology:**

    -   Data Source: AACT (Aggregate Analysis of ClinicalTrials.gov).

    -   Feature Engineering: Regex-based classification on `brief_title` and `intervention_model`.

    -   QA/QC: Filtered for incomplete observations (years \> 2026) and low-sample-size noise.

## 1. Connect the environment to the database

```{r}
#| label: setup
#| message: false
library(DBI); library(RPostgres); library(tidyverse); library(glue)

con <- dbConnect(
  RPostgres::Postgres(),
  host = Sys.getenv("AACT_DB_HOST"),
  dbname = Sys.getenv("AACT_DB_NAME"),
  user = Sys.getenv("AACT_DB_USER"),
  password = Sys.getenv("AACT_DB_PASS"),
  port = as.integer(Sys.getenv("AACT_DB_PORT")),
  sslmode = "require"
)
```

## 2. Data extraction

```{r}
# Extract NCT IDs related to NSCLC
nsclc_ids <- dbGetQuery(con, "
  SELECT DISTINCT nct_id 
  FROM conditions 
  WHERE downcase_name ~* 'non-small cell lung|nsclc'
")$nct_id

# Batch pull detailed information
nct_list_string <- paste0("'", paste(nsclc_ids, collapse = "','"), "'")

#nsclc_raw <- dbGetQuery(con, paste0("
#  SELECT nct_id, brief_title, phase, start_date, completion_date, overall_status
#  FROM studies
#  WHERE nct_id IN (", nct_list_string, ")
#    AND start_date >= '2010-01-01'
#"))

# Query study information
query <- glue("
  SELECT 
    s.nct_id, 
    s.brief_title, 
    s.phase, 
    s.start_date, 
    d.intervention_model, 
    d.primary_purpose 
  FROM studies s
  LEFT JOIN designs d ON s.nct_id = d.nct_id
  WHERE s.nct_id IN ({nct_list_string}) 
  AND s.start_date >= '2010-01-01'
")

nsclc_raw <- dbGetQuery(con, query)
```

## 3. Data cleaning and standardization

```{r}
nsclc_interventional <- nsclc_raw %>%
  filter(!is.na(phase)) %>%
  mutate(phase_standardized = case_when(
    str_detect(phase, "PHASE2") & str_detect(phase, "PHASE3") ~ "Phase 2/3",
    str_detect(phase, "PHASE1") & str_detect(phase, "PHASE2") ~ "Phase 1/2",
    str_detect(phase, "PHASE2") ~ "Phase 2",
    str_detect(phase, "PHASE3") ~ "Phase 3",
    TRUE ~ "Other"
  )) %>%
  filter(!phase_standardized %in% c("Other", "Phase 1/2"))

# Distribution
table(nsclc_interventional$phase_standardized)
```

## 4. Preliminary Trend analysis

```{r}
#| label: plot-trends
#| fig-cap: "The distribution of two-thirds of NSCLC Phase clinical trials from 2010 to 2025"

# Create figures directory and save the plot
# Ensure the 'figures' directory exists to store exported charts
output_dir <- "figures"
dir.create(output_dir, showWarnings = FALSE)

# Generate the stacked bar chart for clinical trial trends
plot_trends <- nsclc_interventional %>%
  mutate(year = format(as.Date(start_date), "%Y")) %>%
  filter(!is.na(year)) %>%
  group_by(year, phase_standardized) %>%
  summarise(n = n(), .groups = 'drop') %>%
  ggplot(aes(x = year, y = n, fill = phase_standardized)) +
  geom_col(position = "stack") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    x = "Year",
    y = "Trial count",
    fill = "Clinical stage"
  )

# Export the plot to the figures directory
ggsave(
  filename = file.path(output_dir, "Fig. 1 nsclc_trial_trends.png"),
  plot = plot_trends,
  width = 8,
  height = 6,
  dpi = 300
)

plot_trends
```

## 5. Trends of Bayesian Adaptive Designs and Master Protocols

```{r}
# Data acquisition and feature engineering
query <- glue_sql("
  SELECT 
    s.nct_id, s.brief_title, s.phase, s.start_date, s.overall_status, 
    d.intervention_model, d.primary_purpose 
  FROM studies s
  LEFT JOIN designs d ON s.nct_id = d.nct_id
  WHERE s.nct_id IN ({nsclc_ids*}) 
  AND s.start_date >= '2010-01-01'
", .con = con)

nsclc_enriched <- dbGetQuery(con, query) %>%
  mutate(
    year = as.integer(format(as.Date(start_date), "%Y")),
    is_adaptive = str_detect(tolower(brief_title), "bayesian|adaptive|seamless|sequential") | 
                  str_detect(toupper(intervention_model), "SEQUENTIAL|FACTORIAL"),
    is_master = str_detect(tolower(brief_title), "basket|umbrella|platform|master")
  )

# Data preprocessing and filtering
# Exclude data post-2026 and filter out years with insufficient sample sizes
nsclc_filtered <- nsclc_enriched %>%
  filter(year >= 2010, year <= 2026) %>% # Truncate incomplete years
  group_by(year) %>%
  mutate(year_total = n()) %>%           # Calculate total annual sample size
  ungroup() %>%
  filter(year_total > 5)                 # Remove years with n <= 5 (denoising)

# Trend analysis visualization
trend_data <- nsclc_filtered %>%
  group_by(year) %>%
  summarise(
    adaptive_pct = mean(is_adaptive, na.rm = TRUE) * 100
  )

# Generate the innovation penetration rate bar chart
plot_innovation <- ggplot(trend_data, aes(x = as.factor(year), y = adaptive_pct)) + 
  geom_col(fill = "steelblue") + 
  theme_minimal() +
  labs(
    title = "Innovation Penetration Rate in NSCLC (2010-2026)", 
    y = "Percentage (%)", 
    x = "Year"
  )

# Export the plot to the figures directory
ggsave(
  filename = file.path(output_dir, "Fig. 2 nsclc_innovation_penetration.png"),
  plot = plot_innovation,
  width = 8,
  height = 6,
  dpi = 300
)

plot_innovation

# Comparative success rate analysis
success_analysis <- nsclc_filtered %>%
  group_by(is_adaptive) %>%
  summarise(
    total = n(),
    success_rate = sum(overall_status == "COMPLETED", na.rm = TRUE) / n()
  )

success_analysis
```

### **Key Findings:**

-   The analysis shows that Phase II clinical trials consistently accounted for the largest proportion of NSCLC studies between 2010 and 2026, indicating that early-stage efficacy evaluation remains a major focus of drug development.

-   The proportion of adaptive trial designs increased over time, suggesting a gradual shift from traditional fixed designs to more flexible clinical trial strategies. This trend reflects the growing use of innovative trial methodologies in NSCLC research.

-   A simple comparison also showed that adaptive trials had a lower completion rate than conventional trials. One possible explanation is that adaptive designs are more likely to stop early when interim results indicate limited benefit, rather than continuing to study completion.

-   Overall, this exploratory analysis demonstrates changing clinical trial practices in NSCLC and provides a foundation for future analyses of trial design and development efficiency.

# **Level 2:** Clinical Efficacy & Development Efficiency Analysis

-   **Objective:** To evaluate the impact of trial design strategies (traditional vs. adaptive/master protocols) on development cycles and clinical endpoint selection, assessing their contribution to overall R&D efficiency.

-   **Methodology:**

    -   **Data Engineering:** Batch extraction via Glue; standardization of design features and clinical metrics.

    -   **Feature Normalization:** Conversion of unstructured text into quantitative metrics (e.g., `duration_months`).

    -   **Statistical Approach:** Comparative analysis using boxplots and the Wilcoxon rank-sum test to evaluate the influence of trial design on study duration.

## 1. Data Extraction and Preprocessing

```{r}
library(purrr)

# Query study information
nct_list_full <- paste0("'", paste(nsclc_ids, collapse = "','"), "'")

query_studies <- glue("
  SELECT s.nct_id, s.brief_title, s.phase, s.start_date, s.completion_date, 
         d.intervention_model, d.primary_purpose
  FROM studies s 
  LEFT JOIN designs d ON s.nct_id = d.nct_id 
  WHERE s.nct_id IN ({nct_list_full}) 
  AND s.start_date >= '2010-01-01' 
")
nsclc_enriched <- dbGetQuery(con, query_studies)

# Split NCT IDs into chunks for batch processing
# Define chunk size (e.g., 500 IDs per batch)
chunk_size <- 500
nct_chunks <- split(nsclc_ids, ceiling(seq_along(nsclc_ids) / chunk_size))

# Extract outcome information via batch processing
all_outcomes_list <- list()

for (i in seq_along(nct_chunks)) {
  current_chunk <- nct_chunks[[i]]
  chunk_string <- paste0("'", paste(current_chunk, collapse = "','"), "'")
  
  query_outcomes <- glue("
    SELECT nct_id, outcome_type, title, time_frame, description
    FROM outcomes
    WHERE nct_id IN ({chunk_string})
  ")
  
  all_outcomes_list[[i]] <- dbGetQuery(con, query_outcomes)
  message(paste("Processed batch", i, "of", length(nct_chunks)))
}

# Merge results
nsclc_outcomes <- bind_rows(all_outcomes_list)

# Identify core endpoints
nsclc_outcomes_clean <- nsclc_outcomes %>%
  mutate(
    endpoint_category = case_when(
      str_detect(tolower(title), "progression-free|pfs") ~ "PFS",
      str_detect(tolower(title), "overall survival|os") ~ "OS",
      str_detect(tolower(title), "response rate|orr") ~ "ORR",
      TRUE ~ "Other"
    )
  ) %>%
  filter(endpoint_category != "Other")

# Link back to the original trial dataset
# Ensure the 'is_adaptive' column exists in nsclc_enriched before joining
nsclc_enriched <- nsclc_enriched %>%
  mutate(
    is_adaptive = str_detect(tolower(brief_title), "bayesian|adaptive|seamless|sequential") | 
                  str_detect(toupper(intervention_model), "SEQUENTIAL|FACTORIAL")
  )

# Perform final join
analysis_df <- nsclc_enriched %>%
  left_join(nsclc_outcomes_clean, by = "nct_id")
```

## 2. Trial Duration and Efficiency Analysis

```{r}
# Calculate development duration (in months)
analysis_efficiency <- analysis_df %>%
  mutate(
    start = as.Date(start_date),
    end = as.Date(completion_date),
    duration_months = as.numeric(end - start) / 30.44
  ) %>%
  filter(duration_months > 0 & duration_months < 120) # Remove outliers

# Compare development duration between adaptive and traditional designs
plot_duration <- ggplot(analysis_efficiency, aes(x = as.factor(is_adaptive), y = duration_months, fill = as.factor(is_adaptive))) + 
  geom_boxplot() +
  labs(
    title = "Clinical Trial Duration: Adaptive vs. Traditional", 
    x = "Is Adaptive Design", 
    y = "Duration (Months)"
  ) +
  theme_minimal()

# Export the plot to the figures directory
ggsave(
  filename = file.path(output_dir, "Fig. 3 nsclc_trial_duration_comparison.png"),
  plot = plot_duration,
  width = 8,
  height = 6,
  dpi = 300
)

plot_duration

# Comparison after removing NA values
clean_data <- analysis_efficiency %>% filter(!is.na(is_adaptive))
wilcox.test(duration_months ~ is_adaptive, data = clean_data)
```

## 3. Endpoint Strategy by Clinical Phase

```{r}
# Standardize clinical trial phases
# Re-standardize the phase column to ensure consistency across the dataset
analysis_df <- analysis_df %>%
  mutate(
    phase_standardized = case_when(
      str_detect(phase, "PHASE2") & str_detect(phase, "PHASE3") ~ "Phase 2/3",
      str_detect(phase, "PHASE1") & str_detect(phase, "PHASE2") ~ "Phase 1/2",
      str_detect(phase, "PHASE2") ~ "Phase 2",
      str_detect(phase, "PHASE3") ~ "Phase 3",
      TRUE ~ "Other"
    )
  )

# Calculate and display the distribution of primary endpoints
# Join trial metadata with outcomes and compute the percentage of each endpoint by phase
endpoint_strategy <- nsclc_outcomes_clean %>%
  left_join(analysis_df %>% select(nct_id, phase_standardized), by = "nct_id") %>%
  filter(!is.na(phase_standardized) & phase_standardized != "Other") %>%
  group_by(phase_standardized, endpoint_category) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(phase_standardized) %>%
  mutate(percent = count / sum(count) * 100)

knitr::kable(
  endpoint_strategy %>% arrange(phase_standardized, desc(percent)),
  caption = "Distribution of Primary Endpoints by Clinical Trial Phase",
  digits = 1
)
```

### Key Findings

-   **Efficiency Metrics:** Statistical analysis (p\>0.05) reveals no significant difference in trial duration between adaptive designs (via keyword matching) and traditional methods, suggesting that design complexity does not inherently accelerate development speed. The shorter observed durations in "NA" categories are primarily attributed to small-scale or non-interventional studies.

-   **Strategic Endpoint Hierarchy:** A distinct hierarchy is evident in clinical development: early-phase trials emphasize safety and rapid signals (ORR/PFS), whereas Phase III trials shift toward Overall Survival (OS) as the definitive measure of utility. The prevalence of OS in Phase I/II trials suggests a cautious, long-term monitoring strategy for molecularly targeted sub-cohorts.

# Level 3: Biological Rationale

-   **Objective:** To provide a genomic context for clinical findings by analyzing somatic mutation profiles in Lung Adenocarcinoma (LUAD).

-   **Methodology:**

    -   Data acquisition from TCGA-LUAD via `TCGAbiolinks`.

    -   Variant analysis and summarization using `maftools` to correlate clinical trial focus with underlying molecular landscapes.

```{r}
# Package installation and loading
# if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install("TCGAbiolinks")
library(TCGAbiolinks)
library(SummarizedExperiment)

# Define query for Lung Adenocarcinoma (LUAD)
query <- GDCquery(
  project = "TCGA-LUAD",
  data.category = "Simple Nucleotide Variation",
  data.type = "Masked Somatic Mutation",
  workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)

# Download data
# GDCdownload(query)

# Prepare and load data
maf_data <- GDCprepare(query)

# Visualization of mutation landscape
library(maftools)
maf_obj <- read.maf(maf = maf_data)
plotmafSummary(maf = maf_obj, rmOutlier = TRUE, addStat = 'median', dashboard = TRUE)

# Define the file path for the mutation summary plot
maf_plot_path <- file.path(output_dir, "Fig. 4 nsclc_mutation_summary.png")

# Generate and save the maftools dashboard
# We use the png() device to ensure the dashboard layout is captured as an image file
png(maf_plot_path, width = 1200, height = 800, res = 120)
plotmafSummary(
  maf = maf_obj, 
  rmOutlier = TRUE, 
  addStat = 'median', 
  dashboard = TRUE
)
dev.off() # Close the graphical device

```

### Key Findings

-   **Genomic Heterogeneity**: Wide variance in mutational burden (median: 158.5 variants/sample) renders "one-size-fits-all" trials inefficient.

-   **Driver Identification**: High-frequency mutations (e.g., *TP53* 50%, *KRAS* 26%) confirm NSCLC as a collection of mutation-specific sub-diseases.

-   **Rationale for Master Protocols**: These findings support Bayesian adaptive designs and Master Protocols to efficiently match patients to targeted therapies within fragmented sub-cohorts.

**AI Collaboration Statement:** The author utilized Gemini, an AI-powered collaborator, to assist with code optimization, statistical logic verification, and technical documentation drafting throughout this project. All scientific methodologies and analytical conclusions remain the responsibility of the author.
